---
title: "PaleoTrees WCE"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: fill
    theme: spacelab
    self_contained: false
    lib_dir: libs
    navbar:
      - { title: "GitHub", href: "https://github.com/", align: right, icon: "fa-github" }
---

<style>
/* ============================================================
   PALEOTREES WCE DASHBOARD — STYLESHEET
   Design: Academic / scientific journal quality
   ============================================================ */

/* --- Fonts --- */
@import url('https://fonts.googleapis.com/css2?family=Inter:ital,wght@0,300;0,400;0,500;0,600;0,700;1,400&display=swap');

/* --- Root Design Tokens --- */
:root {
  --navy-900:          #0f172a;
  --navy-800:          #1e293b;
  --navy-700:          #334155;
  --slate-400:         #94a3b8;
  --slate-300:         #cbd5e1;
  --slate-200:         #e2e8f0;
  --slate-100:         #f1f5f9;
  --body-bg:           #fafaf8;
  --card-bg:           #ffffff;
  --text-base:         #1e293b;
  --text-muted:        #64748b;
  --accent-amber:      #d97706;
  --accent-teal:       #0e7490;
  --accent-teal-light: #e0f2fe;
  --radius:            10px;
  --font-base:         'Inter', -apple-system, BlinkMacSystemFont, 'Segoe UI', sans-serif;
}

/* --- Base Body --- */
body {
  font-family: var(--font-base) !important;
  background-color: var(--body-bg);
  color: var(--text-base);
  font-size: 14px;
  line-height: 1.6;
  -webkit-font-smoothing: antialiased;
  -moz-osx-font-smoothing: grayscale;
}

/* --- Navbar: Dark Gray --- */
.navbar, .navbar-inverse {
  background-color: #575757 !important;
  background-image: none !important;
  border-bottom: none !important;
  border-color: #575757 !important;
  box-shadow: 0 1px 3px rgba(0,0,0,0.3), 0 1px 2px rgba(0,0,0,0.2);
  min-height: 56px;
}
.navbar-brand {
  font-family: var(--font-base);
  font-weight: 700;
  font-size: 1.8rem;
  color: #ffffff !important;
  letter-spacing: 0.02em;
  padding-top: 10px;
  text-transform: none;
}
.navbar-inverse .navbar-nav > li > a {
  color: var(--slate-300) !important;
  font-family: var(--font-base);
  font-weight: 500;
  font-size: 1.3rem;
  letter-spacing: 0.01em;
  padding-top: 14px;
  padding-bottom: 16px;
  transition: color 0.15s ease;
}
.navbar-inverse .navbar-nav > li > a:hover {
  color: #ffffff !important;
  background-color: rgba(255,255,255,0.07) !important;
}
/* Active tab: amber underline */
.navbar-inverse .navbar-nav > .active > a,
.navbar-inverse .navbar-nav > .active > a:hover,
.navbar-inverse .navbar-nav > .active > a:focus {
  color: #ffffff !important;
  background-color: transparent !important;
  border-bottom: 3px solid var(--accent-amber) !important;
  font-weight: 600;
}

/* --- Dashboard Card Wrapper --- */
.chart-wrapper {
  background: var(--card-bg);
  border: 1px solid var(--slate-200);
  border-left: 4px solid #575757;
  border-radius: var(--radius);
  box-shadow:
    0 1px 3px rgba(0,0,0,0.04),
    0 1px 2px rgba(0,0,0,0.03);
  padding: 18px 20px 16px 20px;
  margin-bottom: 20px;
  transition: box-shadow 0.2s ease;
}
.chart-wrapper:hover {
  box-shadow:
    0 4px 12px rgba(0,0,0,0.07),
    0 2px 4px rgba(0,0,0,0.04);
}

/* --- Chart / Panel Titles --- */
.chart-title {
  font-family: var(--font-base);
  font-weight: 600;
  font-size: 1.3rem;
  color: var(--navy-800);
  letter-spacing: -0.01em;
  border-bottom: 1px solid var(--slate-200);
  padding-bottom: 10px;
  margin-bottom: 14px;
  line-height: 1.3;
}

/* ============================================================
   DATATABLES OVERRIDES
   ============================================================ */

/* Extra scroll room so the last rows aren't clipped */
.dataTables_scrollBody table.dataTable {
  margin-bottom: 80px !important;
}

.dataTables_wrapper .dataTables_length,
.dataTables_wrapper .dataTables_filter,
.dataTables_wrapper .dataTables_info,
.dataTables_wrapper .dataTables_processing,
.dataTables_wrapper .dataTables_paginate {
  font-family: var(--font-base);
  color: var(--text-muted);
  font-size: 0.8rem;
  font-weight: 400;
  padding: 4px 0;
}

/* Search input */
.dataTables_wrapper .dataTables_filter input {
  border: 1px solid var(--slate-300);
  border-radius: 6px;
  padding: 4px 10px;
  font-size: 0.8rem;
  font-family: var(--font-base);
  outline: none;
  transition: border-color 0.15s ease, box-shadow 0.15s ease;
}
.dataTables_wrapper .dataTables_filter input:focus {
  border-color: var(--accent-teal);
  box-shadow: 0 0 0 3px rgba(14,116,144,0.12);
}

/* Length select */
.dataTables_wrapper .dataTables_length select {
  border: 1px solid var(--slate-300);
  border-radius: 6px;
  padding: 3px 8px;
  font-size: 0.8rem;
  font-family: var(--font-base);
}

/* Column filter row inputs */
.dataTables_wrapper thead tr.dt-filterRow input,
.dataTables_wrapper thead tr.dt-filterRow select {
  border: 1px solid var(--slate-300);
  border-radius: 5px;
  padding: 3px 6px;
  font-size: 0.75rem;
  font-family: var(--font-base);
  width: 100%;
  box-sizing: border-box;
}
.dataTables_wrapper thead tr.dt-filterRow input:focus,
.dataTables_wrapper thead tr.dt-filterRow select:focus {
  outline: none;
  border-color: var(--accent-teal);
}

/* Table header cells */
table.dataTable thead th,
table.dataTable thead td {
  font-family: var(--font-base);
  font-weight: 600;
  font-size: 0.95rem;
  color: var(--navy-800);
  letter-spacing: 0.04em;
  text-transform: uppercase;
  background-color: var(--slate-100);
  border-bottom: 2px solid var(--slate-300) !important;
  padding: 9px 12px;
  white-space: nowrap;
}

/* Table body rows */
table.dataTable tbody tr {
  font-size: 0.8rem;
  font-family: var(--font-base);
  transition: background-color 0.1s ease;
}
table.dataTable tbody tr:hover {
  background-color: var(--accent-teal-light) !important;
}
table.dataTable tbody tr.odd  { background-color: var(--body-bg); }
table.dataTable tbody tr.even { background-color: #ffffff; }
table.dataTable tbody td {
  padding: 7px 12px;
  border-top: 1px solid var(--slate-100);
  color: var(--text-base);
  vertical-align: middle;
}
table.dataTable.no-footer {
  border-bottom: 2px solid var(--slate-200);
}

/* Pagination buttons */
.dataTables_wrapper .dataTables_paginate .paginate_button {
  font-family: var(--font-base) !important;
  font-size: 0.8rem !important;
  border-radius: 5px !important;
  padding: 3px 8px !important;
}
.dataTables_wrapper .dataTables_paginate .paginate_button.current,
.dataTables_wrapper .dataTables_paginate .paginate_button.current:hover {
  background: var(--navy-800) !important;
  color: #ffffff !important;
  border-color: var(--navy-800) !important;
}
.dataTables_wrapper .dataTables_paginate .paginate_button:hover {
  background: var(--slate-100) !important;
  color: var(--navy-800) !important;
  border-color: var(--slate-300) !important;
}

/* ============================================================
   SCROLLBAR (Webkit: Chrome, Safari, Edge)
   ============================================================ */
::-webkit-scrollbar           { width: 7px; height: 7px; }
::-webkit-scrollbar-track     { background: var(--slate-100); border-radius: 4px; }
::-webkit-scrollbar-thumb     { background: var(--slate-400); border-radius: 4px; }
::-webkit-scrollbar-thumb:hover { background: var(--navy-700); }
.dataTables_scrollBody::-webkit-scrollbar        { width: 6px; height: 6px; }
.dataTables_scrollBody::-webkit-scrollbar-thumb  { background: var(--slate-400); border-radius: 3px; }

/* ============================================================
   PLOTLY STYLING
   ============================================================ */
.plotly .hovertext text,
.plotly .hovertext {
  font-family: var(--font-base) !important;
  font-size: 12px !important;
}
.modebar                    { background: transparent !important; }
.modebar-btn path           { fill: var(--slate-400) !important; }
.modebar-btn:hover path     { fill: var(--navy-800) !important; }

/* ============================================================
   RESPONSIVE
   ============================================================ */
/* --- Download Button --- */
.btn-download {
  display: inline-flex;
  align-items: center;
  gap: 6px;
  background-color: #575757;
  color: #ffffff;
  font-family: Inter, sans-serif;
  font-weight: 500;
  font-size: 12px;
  padding: 9px 20px;
  border: none;
  border-radius: 6px;
  cursor: pointer;
  transition: background-color 0.15s ease, box-shadow 0.15s ease;
  text-decoration: none;
}
.btn-download:hover {
  background-color: #6e6e6e;
  box-shadow: 0 2px 8px rgba(0, 0, 0, 0.25);
  color: #ffffff;
  text-decoration: none;
}

/* --- Legend badge for data table --- */
.legend-badge {
  display: inline-block;
  width: 14px;
  height: 14px;
  border-radius: 3px;
  vertical-align: middle;
  margin-right: 5px;
}

@media (max-width: 768px) {
  .chart-wrapper { padding: 12px 14px; border-radius: 8px; }
  .chart-title   { font-size: 0.95rem; }
  .navbar-brand  { font-size: 1rem; }
  /* Prevent plotly from swallowing the page's vertical scroll gesture */
  .js-plotly-plot .svg-container { touch-action: pan-y !important; }
  /* Allow the panel body to scroll so expanded explanations are reachable */
  .chart-stage { overflow-y: auto !important; }
  /* Hide the modebar on mobile — pan and pinch-zoom still work via touch */
  .modebar-container { display: none !important; }
}
</style>

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(here)
library(brms)
library(readxl)
library(DT)
library(terra)
library(reshape2)

# Set up paths relative to the project root
knitr::opts_knit$set(root.dir = here::here())

# ── PaleoTrees shared ggplot2 theme ──────────────────────────────────────────
theme_paleotrees <- function(base_size = 11) {
    theme_minimal(base_size = base_size) %+replace%
        theme(
            plot.background = element_rect(fill = "#ffffff", colour = NA),
            panel.background = element_rect(fill = "#ffffff", colour = NA),
            panel.grid.major.x = element_blank(),
            panel.grid.major.y = element_line(colour = "#e2e8f0", linewidth = 0.4),
            panel.grid.minor = element_blank(),
            axis.title = element_text(size = base_size * 0.9, colour = "#334155", face = "plain"),
            axis.text = element_text(size = base_size * 0.82, colour = "#64748b"),
            axis.text.x = element_text(size = base_size * 0.82, colour = "#64748b", margin = margin(t = 3)),
            axis.text.y = element_text(size = base_size * 0.75, colour = "#64748b", margin = margin(r = 4)),
            axis.ticks = element_line(colour = "#cbd5e1", linewidth = 0.35),
            axis.ticks.length = unit(3, "pt"),
            axis.line = element_blank(),
            legend.title = element_text(size = base_size * 0.88, colour = "#1e293b", face = "bold"),
            legend.text = element_text(size = base_size * 0.82, colour = "#334155"),
            legend.key = element_rect(fill = NA, colour = NA),
            legend.key.size = unit(12, "pt"),
            legend.position = "right",
            legend.background = element_rect(fill = NA, colour = NA),
            plot.title = element_text(size = base_size * 1.15, colour = "#0f172a", face = "bold", hjust = 0, margin = margin(b = 6)),
            plot.subtitle = element_text(size = base_size * 0.92, colour = "#64748b", hjust = 0, margin = margin(b = 10)),
            plot.margin = margin(t = 8, r = 12, b = 8, l = 8)
        )
}
```

```{r global_data, include=FALSE}
# 1. Load common datasets
dataset_final <- read_xlsx(here("data/data_1/dataset_final_12_02_26.xlsx"))
native_genera <- read.csv(here("data/data_1/native_genera.csv"))[, -1]
native_species <- read.csv(here("data/data_1/native_species.csv"))[, -1]
trees_final <- read.csv(here("data/data_1/trees_final.csv"))

# 2. Colors
colors_periods <- c(
    "Holocene" = "#DCE8E9",
    "Eem" = "#CBF5FC",
    "Holstein" = "#B4E2E3",
    "Cromer" = "#99D9DD",
    "Waal" = "#7FC5DD",
    "Tegelen" = "#4FB3D4",
    "Pleistocene" = "#99d9dd",
    "Pliocene" = "#FFF6B2",
    "Miocene" = "#FFEB3D",
    "Oligocene" = "#FFC694",
    "Eocene" = "#FFBC87",
    "Paleocene" = "#FB9A66",
    "Unknown" = "#cccccc"
)
cat_colors <- c("Extant genera" = "#FFD579", "Holocene regional tree flora" = "#237B55", "Regionally extinct genera" = "#69A0C9")

# ── Shared Plotly style objects ───────────────────────────────────────────────
plotly_font <- list(
    family = "Inter, -apple-system, BlinkMacSystemFont, sans-serif",
    size   = 12,
    color  = "#334155"
)
plotly_axis_style <- list(
    tickfont  = list(family = "Inter, sans-serif", size = 11, color = "#64748b"),
    titlefont = list(family = "Inter, sans-serif", size = 12, color = "#334155"),
    gridcolor = "#e2e8f0",
    gridwidth = 0.5,
    linecolor = "#cbd5e1",
    linewidth = 1,
    zeroline  = FALSE,
    showline  = FALSE
)
plotly_layout_base <- list(
    font = plotly_font,
    paper_bgcolor = "#ffffff",
    plot_bgcolor = "#ffffff",
    hoverlabel = list(
        bgcolor = "#1e293b",
        bordercolor = "#0f172a",
        font = list(family = "Inter, sans-serif", size = 12, color = "#ffffff")
    ),
    legend = list(
        font        = list(family = "Inter, sans-serif", size = 11, color = "#334155"),
        bgcolor     = "rgba(255,255,255,0.92)",
        bordercolor = "#e2e8f0",
        borderwidth = 1
    )
)
```

Niche Models {data-icon="fa-chart-area"}
=======================================================================

Row
-----------------------------------------------------------------------

### Climatic niche centers

<details style="margin-bottom: 10px; font-size: 12px; font-family: 'Inter', sans-serif;">
<summary style="cursor: pointer; font-weight: 500; color: #3b82f6; display: inline-block; padding: 4px 8px; background: #eff6ff; border-radius: 4px; border: 1px solid #bfdbfe;"><i class="fa fa-info-circle"></i> Explanation</summary>
<div style="margin-top: 8px; padding: 10px; background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 4px; color: #475569; line-height: 1.5; max-height: 55vh; overflow-y: auto;">
**Average locations in climate space of Cenozoic tree genera from Western and Central Europe.** Estimated climatic centers of 200 tree genera, comparing regionally extinct genera (blue) and extant genera (yellow). Climatic centers of the Holocene regional tree flora (green), comprising the same genera as in extant genera but modelled using only native species, were also estimated. Hover over a genus to find more information. The Gelman-Rubin criterion (<em>R&#770;</em>) was used to assess Bayesian model convergence, with values closer to 1.0 indicating successful convergence. Current and future climate scenarios for Western and Central Europe have been underlaid, with each square representing a projected 625 km² (25 x 25 km) terrestrial grid cell. Current climate: 1981 – 2010. Moderate warming: SSP1-2.6 (+ 1.3 – 2.4 &deg;C by 2100). Severe warming: SSP5-8.5 (+ 3.3 – 5.7 &deg;C by 2100). Switch to last occurrence to see the timing of the last known fossil occurrence of each genus in Western and Central Europe. Point size scales with the number of accepted tree species within each genus.
<div style="text-align: center; margin-top: 12px;"><img src="stratigraphy.svg" alt="Stratigraphic chart" style="max-width: 80%; height: auto;"></div></div>
</details>


```{r centers_data, include=FALSE}
# Load niche center models
directory_path <- here("data/niche_centers")
file_paths <- list.files(directory_path, full.names = TRUE)

all_data_centers <- data.frame()
if (length(file_paths) > 0) {
    model_list <- lapply(file_paths, readRDS)
    model_names <- sub(".*model_(.*)\\.RData", "\\1", basename(file_paths))
    genus_names <- sub("_na$", "", model_names)

    for (g in 1:length(genus_names)) {
        tryCatch(
            {
                genus_name <- genus_names[[g]]
                model_name <- model_names[[g]]
                model <- model_list[[g]]
                Rhat <- tryCatch(max(rhat(model)), error = function(e) NA)
                species_n <- max(1, length(unique(model$data$species)))
                para <- as_draws_df(model)
                if (all(c("b_bio1_Intercept", "b_bio12_Intercept") %in% names(para))) {
                    para_means <- colMeans(select(para, b_bio1_Intercept, b_bio12_Intercept))
                    df <- data.frame(genus = genus_name, model = model_name, bio1 = para_means[1], bio12 = para_means[2], Rhat = Rhat, n = species_n)
                    all_data_centers <- rbind(all_data_centers, df)
                }
            },
            error = function(e) {
                return(NULL)
            }
        )
    }
}

if (nrow(all_data_centers) > 0) {
    all_data_centers$category <- ifelse(all_data_centers$genus %in% native_species | grepl("_na$", all_data_centers$model), "Holocene regional tree flora",
        ifelse(all_data_centers$genus %in% native_genera, "Extant genera", "Regionally extinct genera")
    )

    paleotrees_adj <- dataset_final
    paleotrees_adj$last_presence <- apply(paleotrees_adj[, -1], 1, function(x) {
        if (any(x == 1)) names(x)[max(which(x == 1))] else NA
    })

    all_data_centers <- merge(all_data_centers, paleotrees_adj[, c(1, 13)], by = "genus")
    all_data_centers$last_presence_group <- ifelse(all_data_centers$last_presence %in% c("Eem", "Tegelen", "Waal", "Cromer", "Holstein"), "Pleistocene", all_data_centers$last_presence)

    n_max <- max(all_data_centers$n, na.rm = TRUE)
    all_data_centers$scaled_size <- 7 + 15 * sqrt(all_data_centers$n / n_max)

    all_data_centers$hover_text <- paste(
        "<b>Genus:</b>", all_data_centers$genus,
        "<br><b>Status:</b>", all_data_centers$category,
        "<br><b>Tree species count:</b>", all_data_centers$n,
        "<br><b>Last occurrence:</b>", all_data_centers$last_presence,
        "<br><b>MAT:</b>", round(all_data_centers$bio1, 2), " &deg;C",
        "<br><b>AP:</b>", round(all_data_centers$bio12, 0), " mm",
        "<br><b>R&#770;:</b>", sprintf("%.2f", all_data_centers$Rhat)
    )
}

bg_data <- data.frame()
try(
    {
        process_raster_robust <- function(path, label) {
            if (file.exists(path)) {
                r <- rast(path)
                b1_idx <- grep("bio1_", names(r))
                if (length(b1_idx) == 0) b1_idx <- grep("bio1$", names(r))
                b12_idx <- grep("bio12_", names(r))
                if (length(b12_idx) == 0) b12_idx <- grep("bio12$", names(r))
                b5_idx <- grep("bio5_", names(r))
                if (length(b5_idx) == 0) b5_idx <- grep("bio5$", names(r))
                b6_idx <- grep("bio6_", names(r))
                if (length(b6_idx) == 0) b6_idx <- grep("bio6$", names(r))

                if (length(b1_idx) > 0 && length(b12_idx) > 0 && length(b5_idx) > 0 && length(b6_idx) > 0) {
                    r_sub <- r[[c(b1_idx[1], b12_idx[1], b5_idx[1], b6_idx[1])]]
                    # Equal-area projection (LAEA Europe) to match analysis code (EPSG:3035)
                    laea_proj <- "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"
                    r_sub <- tryCatch(terra::project(r_sub, laea_proj), error = function(e) r_sub)
                    df <- as.data.frame(r_sub, xy = TRUE)
                    colnames(df) <- c("x", "y", "bio1", "bio12", "bio5", "bio6")
                    df <- df %>% mutate(group = label)
                    return(df)
                }
            }
            return(NULL)
        }
        bg_data <- rbind(
            process_raster_robust(here("data/data_2/bioclim_current_0_25deg.tif"), "Current"),
            process_raster_robust(here("data/data_2/bioclim_ssp126_0_25deg.tif"), "SSP126"),
            process_raster_robust(here("data/data_2/bioclim_ssp585_0_25deg.tif"), "SSP585")
        ) %>% na.omit()
    },
    silent = TRUE
)
```

```{r plot_centers}
if (exists("all_data_centers") && nrow(all_data_centers) > 0) {
    plot_data <- all_data_centers %>% filter((is.na(Rhat) | Rhat <= 1.05))
    # Widened filters to ensure Paleocene (potentially warmer/wetter) is included
    plot_data <- plot_data %>% filter(bio1 >= 0 & bio1 <= 35 & bio12 >= 400 & bio12 <= 3000)

    fig <- plot_ly() # Let flexdashboard handle height in 'fill' mode

    if (nrow(bg_data) > 0) {
        scenarios <- unique(bg_data$group)
        cols_scen <- c("Current" = "#909090", "SSP126" = "#FF9538", "SSP585" = "#EB0000")

        hex_to_rgba <- function(hex, alpha) {
            r <- strtoi(substr(hex, 2, 3), 16L)
            g <- strtoi(substr(hex, 4, 5), 16L)
            b <- strtoi(substr(hex, 6, 7), 16L)
            sprintf("rgba(%d,%d,%d,%.2f)", r, g, b, alpha)
        }

        for (scen in scenarios) {
            sub <- bg_data[bg_data$group == scen, ] %>% filter(bio1 >= 0 & bio1 <= 35 & bio12 >= 400 & bio12 <= 3000)
            col <- cols_scen[scen]
            cs <- list(list(0, hex_to_rgba(col, 0)), list(1, hex_to_rgba(col, 0.65)))
            kde <- MASS::kde2d(sub$bio1, sub$bio12, n = 100, lims = c(0, 35, 400, 3000))
            kde$z <- kde$z / max(kde$z, na.rm = TRUE) # normalise like ndensity
            fig <- fig %>% add_trace(
                x = kde$x, y = kde$y, z = t(kde$z),
                type = "contour",
                colorscale = cs, showscale = FALSE,
                contours = list(start = 1 / 7, end = 1, size = 1 / 7, coloring = "fill", showlines = FALSE),
                name = scen, visible = TRUE, showlegend = TRUE, hoverinfo = "none"
            )
        }
    }

    marker_defaults <- list(sizemode = "diameter", line = list(width = 1, color = "black"))

    plot_data$symbol_shape <- ifelse(plot_data$category == "Holocene regional tree flora", "triangle-up", "circle")

    cats <- unique(plot_data$category)
    for (cat in cats) {
        sub <- plot_data[plot_data$category == cat, ]
        marker_symbol <- unique(sub$symbol_shape) # Since we filter by category, this is uniform

        fig <- fig %>% add_trace(
            data = sub, x = ~bio1, y = ~bio12, text = ~hover_text, hoverinfo = "text",
            type = "scatter", mode = "markers", name = cat,
            marker = c(list(color = cat_colors[cat], size = sub$scaled_size, opacity = 0.9, symbol = marker_symbol), marker_defaults),
            visible = TRUE
        )
    }

    lps <- unique(plot_data$last_presence_group)
    # Order chronologically (names(colors_periods) is defined order: Holocene -> Paleocene)
    lps <- lps[order(match(lps, names(colors_periods)))]

    for (lp in lps) {
        sub <- plot_data[plot_data$last_presence_group == lp, ]
        col <- colors_periods[lp]
        if (is.na(col)) col <- "gray"
        fig <- fig %>% add_trace(
            data = sub, x = ~bio1, y = ~bio12, text = ~hover_text, hoverinfo = "text",
            type = "scatter", mode = "markers", name = lp,
            marker = c(list(color = col, size = sub$scaled_size, opacity = 0.9, symbol = ~symbol_shape), marker_defaults),
            visible = FALSE
        )
    }

    # Strict Limits, No Scaleanchor (FIX for Iteration 19)
    fig <- fig %>%
        layout(
            font = plotly_font,
            paper_bgcolor = "#ffffff",
            plot_bgcolor = "#ffffff",
            hoverlabel = plotly_layout_base$hoverlabel,
            xaxis = modifyList(plotly_axis_style, list(
                title = list(
                    text = "Mean annual temperature [\u00B0C]",
                    font = list(size = 15, color = "#334155"),
                    standoff = 20
                ),
                showgrid = FALSE,
                zeroline = FALSE,
                range = c(0, 35),
                tickfont = list(size = 11),
                constrain = "domain",
                showline = TRUE,
                linewidth = 1,
                linecolor = "black",
                mirror = TRUE,
                ticks = "outside",
                tickcolor = "black"
            )),
            yaxis = modifyList(plotly_axis_style, list(
                title = list(
                    text = "Annual precipitation [mm]",
                    font = list(size = 15, color = "#334155")
                ),
                showgrid = FALSE,
                gridcolor = "#e2e8f0",
                zeroline = FALSE,
                range = c(400, 3000),
                tickfont = list(size = 11),
                constrain = "domain",
                showline = TRUE,
                linewidth = 1,
                linecolor = "black",
                mirror = TRUE,
                ticks = "outside",
                tickcolor = "black"
            )),
            legend = list(
                x           = 0.02,
                y           = 0.98,
                xanchor     = "left",
                font        = list(family = "Inter, sans-serif", size = 11, color = "#334155"),
                bgcolor     = "rgba(255,255,255,0.8)",
                bordercolor = "#e2e8f0",
                borderwidth = 1
            ),
            dragmode = "pan",
            uirevision = "constant",
            margin = list(l = 60, r = 20, t = 40, b = 100)
        ) %>%
        config(displayModeBar = FALSE, scrollZoom = TRUE) %>%
        htmlwidgets::onRender(sprintf("
            function(el) {
                var nCats = %d;
                var nLps  = %d;

                /* -- Status / Last occurrence toggle buttons -- */
                var btnStyle = 'font-family:Inter,sans-serif;font-size:12px;color:#1e293b;border:1px solid #cbd5e1;border-radius:3px;padding:4px 14px;cursor:pointer;line-height:1.2;pointer-events:auto;';

                var btnStatus = document.createElement('button');
                btnStatus.innerHTML = 'Status';
                btnStatus.style.cssText = btnStyle + 'background:#cbd5e1;';

                var btnLast = document.createElement('button');
                btnLast.innerHTML = 'Last occurrence';
                btnLast.style.cssText = btnStyle + 'background:#f1f5f9;';

                var btnWrap = document.createElement('div');
                btnWrap.style.cssText = 'position:absolute;z-index:1000;top:6px;left:0;right:0;display:flex;justify-content:center;gap:6px;visibility:hidden;pointer-events:none;';
                btnWrap.appendChild(btnStatus);
                btnWrap.appendChild(btnLast);

                /* -- Toggle Legend button -- */
                var legendVisible = true;
                var btnLegend = document.createElement('button');
                btnLegend.textContent = 'Toggle Legend';
                btnLegend.style.cssText = [
                    'position:absolute',
                    'z-index:1000',
                    'font-family:Inter,sans-serif',
                    'font-size:12px',
                    'color:#1e293b',
                    'background:#f1f5f9',
                    'border:1px solid #cbd5e1',
                    'border-radius:3px',
                    'padding:4px 8px',
                    'cursor:pointer',
                    'line-height:1.2',
                    'visibility:hidden'
                ].join(';');

                function pinButtons() {
                    var layout = el._fullLayout;
                    if (!layout) return;
                    var mt = layout.margin.t;
                    var ml = layout.margin.l || 0;
                    var mr = layout.margin.r || 0;
                    btnWrap.style.paddingLeft  = ml + 'px';
                    btnWrap.style.paddingRight = mr + 'px';
                    btnWrap.style.visibility = 'visible';
                    btnLegend.style.top   = mt + 'px';
                    btnLegend.style.right = mr + 'px';
                    btnLegend.style.visibility = 'visible';
                }

                function setMode(mode) {
                    var vis = [];
                    for (var i = 0; i < 3; i++) vis.push(true);
                    for (var i = 0; i < nCats; i++) vis.push(mode === 'status');
                    for (var i = 0; i < nLps; i++) vis.push(mode !== 'status');
                    Plotly.restyle(el, {visible: vis});
                    btnStatus.style.background = mode === 'status' ? '#cbd5e1' : '#f1f5f9';
                    btnLast.style.background   = mode !== 'status' ? '#cbd5e1' : '#f1f5f9';
                }

                var initialized = false;
                el.addEventListener('plotly_afterplot', function() {
                    pinButtons();
                    if (!initialized) {
                        initialized = true;
                        setMode('status');
                    }
                });
                window.addEventListener('resize', pinButtons);

                btnStatus.addEventListener('click', function() { setMode('status'); });
                btnLast.addEventListener('click',   function() { setMode('last');   });

                btnLegend.addEventListener('click', function() {
                    legendVisible = !legendVisible;
                    Plotly.relayout(el, {showlegend: legendVisible});
                });

                var chartStage = el.closest('.chart-stage') || el.parentNode;
                if (window.getComputedStyle(chartStage).position === 'static') {
                    chartStage.style.position = 'relative';
                }
                chartStage.appendChild(btnWrap);
                el.querySelector('.svg-container').appendChild(btnLegend);
            }
        ", length(cats), length(lps)))

    fig
}
```

### Thermal niche limits

<details style="margin-bottom: 10px; font-size: 12px; font-family: 'Inter', sans-serif;">
<summary style="cursor: pointer; font-weight: 500; color: #3b82f6; display: inline-block; padding: 4px 8px; background: #eff6ff; border-radius: 4px; border: 1px solid #bfdbfe;"><i class="fa fa-info-circle"></i> Explanation</summary>
<div style="margin-top: 8px; padding: 10px; background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 4px; color: #475569; line-height: 1.5; max-height: 55vh; overflow-y: auto;">
**Upper and lower thermal limits of Cenozoic tree genera from Western and Central Europe.** Posteriors of upper and lower thermal limits for 200 tree genera were estimated, comparing regionally extinct genera (blue) and extant genera (yellow). Posteriors of genus-level thermal limits of the Holocene regional tree flora (green) were also estimated, comprising the same genera as in extant genera but modelled using only native species. Models were arranged by increasing upper thermal limit (bottom to top). Color stripes correspond to the last known occurrence of each genus in the fossil record of Western and Central Europe. Hover over a genus to find more information. The Gelman-Rubin criterion (<em>R&#770;</em>) was used to assess Bayesian model convergence, with values closer to 1.0 indicating successful convergence. The distribution of heat and cold extremes (mean daily maximum temperature of the warmest month and mean daily minimum temperature of the coldest month) by land area under current and future climate scenarios in Western and Central Europe is overlaid. Current climate: 1981 – 2010. Moderate warming: SSP1-2.6 (+ 1.3 – 2.4 &deg;C by 2100). Severe warming: SSP5-8.5 (+ 3.3 – 5.7 &deg;C by 2100).
<div style="text-align: center; margin-top: 12px;"><img src="stratigraphy.svg" alt="Stratigraphic chart" style="max-width: 80%; height: auto;"></div></div>
</details>

```{r limits_data_refined, include=FALSE}
directory_path_limits <- here("data/niche_limits")
file_paths_limits <- list.files(directory_path_limits, full.names = TRUE)

all_data_limits_densities <- data.frame()
medians_segments <- data.frame()
ordered_models <- c()

if (length(file_paths_limits) > 0) {
    model_list_limits <- lapply(file_paths_limits, readRDS)
    model_names_limits <- sub(".*model_(.*)\\.RData", "\\1", basename(file_paths_limits))

    medians_df <- data.frame()
    for (g in 1:length(model_list_limits)) {
        try(
            {
                para <- as_draws_df(model_list_limits[[g]])
                if (all(c("b_bio5_Intercept", "b_bio6_Intercept") %in% names(para))) {
                    m5 <- median(para$b_bio5_Intercept)
                    m6 <- median(para$b_bio6_Intercept)
                    medians_df <- rbind(medians_df, data.frame(model = model_names_limits[g], m_bio5 = m5, m_bio6 = m6))
                }
            },
            silent = T
        )
    }
    if (nrow(medians_df) > 0) {
        ordered_models <- medians_df %>%
            arrange(m_bio5) %>%
            pull(model)
    }

    for (g in 1:length(model_list_limits)) {
        tryCatch(
            {
                m_name <- model_names_limits[g]
                model <- model_list_limits[[g]]
                n_species <- tryCatch(max(1, length(unique(model$data$species))), error = function(e) NA)
                para <- as_draws_df(model)
                if (all(c("b_bio5_Intercept", "b_bio6_Intercept") %in% names(para))) {
                    Rhat_val <- tryCatch(max(rhat(model)), error = function(e) NA)
                    d_min <- density(para$b_bio6_Intercept)
                    d_max <- density(para$b_bio5_Intercept)
                    idx <- which(ordered_models == m_name)
                    max_val <- max(c(d_min$y, d_max$y))
                    df_min <- data.frame(x = d_min$x, y_final = (d_min$y / max_val) + idx, model = m_name, Metric = "Min Temp", Rhat = Rhat_val, n = n_species)
                    df_max <- data.frame(x = d_max$x, y_final = (d_max$y / max_val) + idx, model = m_name, Metric = "Max Temp", Rhat = Rhat_val, n = n_species)
                    all_data_limits_densities <- rbind(all_data_limits_densities, df_min, df_max)
                    m5 <- median(para$b_bio5_Intercept)
                    m6 <- median(para$b_bio6_Intercept)
                    medians_segments <- rbind(medians_segments, data.frame(model = m_name, m6 = m6, m5 = m5, y = idx, Rhat = Rhat_val, n = n_species))
                }
            },
            error = function(e) {
                return(NULL)
            }
        )
    }
}
if (nrow(all_data_limits_densities) > 0) {
    all_data_limits_densities$label <- ifelse(grepl("_na$", all_data_limits_densities$model),
        paste0(gsub("_na$", "", all_data_limits_densities$model), "<sub>na</sub>"), all_data_limits_densities$model
    )
    medians_segments$label <- ifelse(grepl("_na$", medians_segments$model),
        paste0(gsub("_na$", "", medians_segments$model), "<sub>na</sub>"), medians_segments$model
    )
    all_data_limits_densities$genus <- gsub("_na$", "", all_data_limits_densities$model)
    all_data_limits_densities$category <- ifelse(all_data_limits_densities$genus %in% native_species | grepl("_na$", all_data_limits_densities$model), "Holocene regional tree flora",
        ifelse(all_data_limits_densities$genus %in% native_genera, "Extant genera", "Regionally extinct genera")
    )
    medians_segments$category <- ifelse(gsub("_na$", "", medians_segments$model) %in% native_species | grepl("_na$", medians_segments$model), "Holocene regional tree flora",
        ifelse(gsub("_na$", "", medians_segments$model) %in% native_genera, "Extant genera", "Regionally extinct genera")
    )
}
```

```{r plot_limits_scrolling}
if (nrow(all_data_limits_densities) > 0) {
    # --- 1. Data Preparation ---
    all_data_limits_densities <- all_data_limits_densities %>%
        group_by(model, Metric) %>%
        arrange(x) %>%
        mutate(
            y_base = min(y_final),
            density_height = round(pmax(0, y_final - y_base) * 2.5, 3)
        ) %>%
        filter(density_height > 0.1 | row_number() %% 8 == 1 | row_number() == 1 | row_number() == n()) %>%
        ungroup()

    all_data_limits_densities$label <- as.character(all_data_limits_densities$label)

    ord_df <- medians_segments %>%
        select(model, label, m5, m6) %>%
        distinct() %>%
        group_by(model, label) %>%
        summarize(m5 = mean(m5), m6 = mean(m6), .groups = "drop")

    # Single sort order: by upper thermal limit (ascending), per the figure description
    ord_min <- ord_df %>%
        arrange(m5) %>%
        pull(model)
    map_y_min <- setNames(seq_along(ord_min), ord_min)

    # --- 2. Calculate Ribbon Heights ---
    all_data_limits_densities$density_height <- pmax(0, all_data_limits_densities$y_final - all_data_limits_densities$y_base)
    all_data_limits_densities$x <- round(all_data_limits_densities$x, 2)

    all_data_limits_densities$y_min_base <- map_y_min[all_data_limits_densities$model]
    all_data_limits_densities$y_min_top <- round(all_data_limits_densities$y_min_base + all_data_limits_densities$density_height, 3)

    medians_segments$y_min <- map_y_min[medians_segments$model]

    # --- 3. Merge Last Presence ---
    if (exists("all_data_centers") && nrow(all_data_centers) > 0) {
        lp_map <- setNames(all_data_centers$last_presence_group, all_data_centers$model)
        all_data_limits_densities$last_presence_group <- lp_map[all_data_limits_densities$model]
        medians_segments$last_presence_group <- lp_map[medians_segments$model]
    }
    all_data_limits_densities$last_presence_group[is.na(all_data_limits_densities$last_presence_group)] <- "Unknown"
    medians_segments$last_presence_group[is.na(medians_segments$last_presence_group)] <- "Unknown"

    # --- 4. Hover Text ---
    all_data_limits_densities$hover_text <- paste0(
        "<b>Genus:</b> ", all_data_limits_densities$genus, "<br>",
        "<b>Tree species count:</b> ", all_data_limits_densities$n, "<br>",
        "<b>Limit:</b> ", all_data_limits_densities$x, " &deg;C<br>",
        "<b>Status:</b> ", all_data_limits_densities$category, "<br>",
        "<b>Last occurrence:</b> ", all_data_limits_densities$last_presence_group, "<br>",
        "<b>R&#770;:</b> ", sprintf("%.2f", all_data_limits_densities$Rhat)
    )
    medians_segments$hover_text <- paste0(
        "<b>Genus:</b> ", gsub("_na$", "", medians_segments$model), "<br>",
        "<b>Tree species count:</b> ", medians_segments$n, "<br>",
        "<b>Center:</b> ", round((medians_segments$m5 + medians_segments$m6) / 2, 2), " &deg;C<br>",
        "<b>Status:</b> ", medians_segments$category, "<br>",
        "<b>Last occurrence:</b> ", medians_segments$last_presence_group, "<br>",
        "<b>R&#770;:</b> ", sprintf("%.2f", medians_segments$Rhat)
    )

    labels_min <- ord_df$label[match(ord_min, ord_df$model)]

    # --- Configuration ---
    common_x_range <- c(-20, 60)
    fixed_margins <- list(l = 100, r = 20)

    # --- Top Plot: Climate Distributions ---
    dens_bio5 <- data.frame()
    dens_bio6 <- data.frame()
    if (nrow(bg_data) > 0) {
        for (scen in c("Current", "SSP126", "SSP585")) {
            if (scen %in% unique(bg_data$group)) {
                d5 <- density(bg_data$bio5[bg_data$group == scen], from = common_x_range[1], to = common_x_range[2], n = 128)
                dens_bio5 <- rbind(dens_bio5, data.frame(x = round(d5$x, 2), y = round(d5$y, 4), group = scen))
                d6 <- density(bg_data$bio6[bg_data$group == scen], from = common_x_range[1], to = common_x_range[2], n = 128)
                dens_bio6 <- rbind(dens_bio6, data.frame(x = round(d6$x, 2), y = round(d6$y, 4), group = scen))
            }
        }
    }

    p_top <- plot_ly(height = 140)
    cols_scen <- c("Current" = "#909090", "SSP126" = "#FF9538", "SSP585" = "#EB0000")
    cols_scen_rgba_fill <- c("Current" = "rgba(144,144,144,0.4)", "SSP126" = "rgba(255,149,56,0.4)", "SSP585" = "rgba(235,0,0,0.4)")
    scen_order <- c("Current", "SSP126", "SSP585")

    for (s in scen_order) {
        if (nrow(dens_bio6) > 0 && s %in% unique(dens_bio6$group)) {
            d6 <- dens_bio6[dens_bio6$group == s, ]
            p_top <- p_top %>% add_trace(
                x = d6$x, y = d6$y, type = "scatter", mode = "lines", fill = "tozeroy",
                line = list(color = cols_scen[s], width = 1.5), fillcolor = cols_scen_rgba_fill[s],
                name = paste("Lower:", s), legendgroup = "Lower", visible = FALSE, hoverinfo = "name+x"
            )
        }
    }
    for (s in scen_order) {
        if (nrow(dens_bio5) > 0 && s %in% unique(dens_bio5$group)) {
            d5 <- dens_bio5[dens_bio5$group == s, ]
            p_top <- p_top %>% add_trace(
                x = d5$x, y = d5$y, type = "scatter", mode = "lines", fill = "tozeroy",
                line = list(color = cols_scen[s], width = 1.5), fillcolor = cols_scen_rgba_fill[s],
                name = paste("Upper:", s), legendgroup = "Upper", visible = TRUE, hoverinfo = "name+x"
            )
        }
    }

    n_s <- length(scen_order)
    p_top <- p_top %>%
        layout(
            xaxis = list(title = "Temperature [\u00B0C]", range = common_x_range, fixedrange = TRUE, side = "bottom", showgrid = FALSE, zeroline = FALSE),
            yaxis = list(showgrid = FALSE, showticklabels = FALSE, title = "", fixedrange = TRUE, zeroline = FALSE),
            margin = c(fixed_margins, list(t = 5, b = 40)),
            showlegend = FALSE
        ) %>%
        config(staticPlot = FALSE, displayModeBar = FALSE) %>%
        htmlwidgets::onRender("function(el) { el.id = 'p-top-climate'; }")

    btn_style_base <- "font-family:Inter,sans-serif;font-size:12px;color:#1e293b;border:1px solid #cbd5e1;border-radius:3px;padding:4px 14px;cursor:pointer;"
    toggle_html <- htmltools::div(
        style = paste0("display:flex;justify-content:center;gap:6px;padding:6px ", fixed_margins$r, "px 4px ", fixed_margins$l, "px;"),
        htmltools::tags$button(
            id = "ptop-btn-lower", style = paste0(btn_style_base, "background:#f1f5f9;"),
            onclick = "ptopToggle('lower')", htmltools::HTML("T<sub>min</sub>")
        ),
        htmltools::tags$button(
            id = "ptop-btn-upper", style = paste0(btn_style_base, "background:#cbd5e1;"),
            onclick = "ptopToggle('upper')", htmltools::HTML("T<sub>max</sub>")
        )
    )


    # --- Main Plot: Niche Limits (single layer: Status coloring) ---
    p_limits <- ggplot() +
        geom_ribbon(
            data = all_data_limits_densities,
            aes(
                x = x, ymin = y_min_base, ymax = y_min_top,
                group = interaction(model, Metric), fill = category, text = hover_text
            ),
            alpha = 0.6
        ) +
        geom_segment(
            data = medians_segments,
            aes(x = m6, xend = m5, y = y_min, yend = y_min, color = category, text = hover_text),
            linewidth = 0.8
        ) +
        scale_fill_manual(values = c(cat_colors, colors_periods)) +
        scale_color_manual(values = c(cat_colors, colors_periods)) +
        scale_x_continuous(limits = common_x_range) +
        scale_y_continuous(limits = c(0.5, nrow(ord_df) + 1.5), expand = c(0, 0)) +
        theme_paleotrees() +
        theme(panel.grid = element_blank(), axis.text.y = element_text(size = 6), legend.position = "none")

    limits_height <- max(400, nrow(ord_df) * 17)
    p_ply <- ggplotly(p_limits, height = limits_height, tooltip = "text")

    # Add last-presence colour squares as always-visible plotly traces
    for (lp in unique(medians_segments$last_presence_group)) {
        sub <- medians_segments[medians_segments$last_presence_group == lp, ]
        col <- colors_periods[lp]
        if (is.na(col)) col <- "#cccccc"
        p_ply <- p_ply %>% add_trace(
            x = rep(50, nrow(sub)), y = sub$y_min,
            type = "scatter", mode = "markers",
            marker = list(symbol = "square", size = 12, color = col, line = list(width = 0)),
            text = paste0(gsub("_na$", "", sub$model), "<br>Last occurrence: ", sub$last_presence_group),
            hoverinfo = "text", showlegend = FALSE, inherit = FALSE
        )
    }

    tickvals <- seq_len(nrow(ord_df))

    p_ply <- p_ply %>% layout(
        dragmode = "pan",
        xaxis = list(
            title = "", showgrid = FALSE, showline = FALSE, showticklabels = FALSE,
            zeroline = FALSE, range = common_x_range, fixedrange = TRUE
        ),
        yaxis = list(
            title = "", showgrid = FALSE, zeroline = FALSE, tickfont = list(size = 11),
            tickmode = "array", tickvals = tickvals, ticktext = labels_min,
            range = c(0.5, nrow(ord_df) + 1.5), fixedrange = TRUE
        ),
        margin = c(fixed_margins, list(t = 20, b = 20))
    )

    # Column headers inside the SVG, visible at the top of the scroll view
    p_ply <- p_ply %>%
        add_annotations(
            x = common_x_range[1], xref = "x",
            y = nrow(ord_df) + 1.0, yref = "y",
            text = "<b>Genus</b>", showarrow = FALSE,
            font = list(size = 12, color = "#334155", family = "Inter, sans-serif"),
            xanchor = "right", yanchor = "middle", xshift = -4
        ) %>%
        add_annotations(
            x = 50, xref = "x",
            y = nrow(ord_df) + 1.0, yref = "y",
            text = "<b>Last occurrence</b>", showarrow = FALSE,
            font = list(size = 12, color = "#334155", family = "Inter, sans-serif"),
            xanchor = "center", yanchor = "middle"
        )

    # Wipe out any shapes to prevent grid lines or previous artifacts
    p_ply$x$layout$shapes <- NULL
    p_ply <- p_ply %>% config(displayModeBar = FALSE, scrollZoom = FALSE)

    # HTML scenario labels — sit between p_top and the scrollable list, wrap on narrow screens
    scen_label_html <- htmltools::div(
        style = paste0("width: 100%; display: flex; flex-wrap: wrap; justify-content: center; gap: 4px 20px; padding: 4px ", fixed_margins$r, "px 4px ", fixed_margins$l, "px; font-family: Inter, sans-serif; font-size: 12px;"),
        htmltools::span(style = paste0("color: ", cols_scen["Current"], "; white-space: nowrap;"), "Current climate in WCE"),
        htmltools::span(style = paste0("color: ", cols_scen["SSP126"], "; white-space: nowrap;"), "Moderate warming SSP1-2.6"),
        htmltools::span(style = paste0("color: ", cols_scen["SSP585"], "; white-space: nowrap;"), "Severe warming SSP5-8.5")
    )

    htmltools::tagList(
        toggle_html,
        scen_label_html,
        htmltools::div(style = "width: 100%; height: 140px; margin-bottom: 0px;", p_top),
        htmltools::div(style = "width: 100%; height: calc(100vh - 420px); overflow-y: auto; touch-action: pan-y;", p_ply)
    )
}
```
Data {data-icon="fa-database"}
=======================================================================

Row
-----------------------------------------------------------------------

### Occurrence Dataset

<details style="margin-bottom: 10px; font-size: 12px; font-family: 'Inter', sans-serif;">
<summary style="cursor: pointer; font-weight: 500; color: #3b82f6; display: inline-block; padding: 4px 8px; background: #eff6ff; border-radius: 4px; border: 1px solid #bfdbfe;"><i class="fa fa-info-circle"></i> Explanation</summary>
<div style="margin-top: 8px; padding: 10px; background: #f8fafc; border: 1px solid #e2e8f0; border-radius: 4px; color: #475569; line-height: 1.5; max-height: 55vh; overflow-y: auto;">
Fossil occurrences of tree genera during the Cenozoic (66 Ma – present) from Western and Central Europe (including Austria, Belgium, Czechia, France, Germany, Ireland, Luxembourg, Netherlands, Poland, Switzerland, and United Kingdom) were compiled using the Paleobiology Database (PBDB) and literature. For the Pleistocene, additional fossil occurrences dated to five temperate stages and interglacial periods were included. Holocene occurrences were obtained from the World Checklist of Vascular Plants (WCVP).
</div>
</details>

```{r data_tab, fig.width=12}
# Load the dashboard dataset
dashboard_data <- read_xlsx(here("data/data_1/dashboard_data_12_02_26.xlsx"))

# Period columns (numeric occurrence data)
period_cols <- c(
    "Paleocene", "Eocene", "Oligocene", "Miocene", "Pliocene",
    "Tegelen", "Waal", "Cromer", "Holstein", "Eem", "Holocene"
)

colors_periods <- c(
    "Paleocene" = "#FF9A5C",
    "Eocene"    = "#FFBC87",
    "Oligocene" = "#FFC694",
    "Miocene"   = "#FFEB3D",
    "Pliocene"  = "#FFF6B2",
    "Tegelen"   = "#4FB3D4",
    "Waal"      = "#7FC5DD",
    "Cromer"    = "#99D9DD",
    "Holstein"  = "#B4E2E3",
    "Eem"       = "#CBF5FC",
    "Holocene"  = "#DCE8E9"
)

# Display labels for values (invisible text — cells are color-only)
dashboard_data_display <- dashboard_data
for (col in period_cols) {
    dashboard_data_display[[col]] <- ifelse(
        dashboard_data[[col]] == 2, "2",
        ifelse(dashboard_data[[col]] == 1, "1", "0")
    )
}

# Build the download button + legend HTML
header_html <- htmltools::div(
    style = "display: flex; justify-content: space-between; align-items: center; margin-bottom: 12px; flex-wrap: wrap; gap: 10px;",
    htmltools::div(
        style = "display: flex; align-items: center; gap: 16px; font-family: Inter, sans-serif; font-size: 12px; color: #475569;",
        htmltools::span(
            htmltools::span(class = "legend-badge", style = "background-color: rgba(30,58,95,0.55);"),
            "PBDB occurrence"
        ),
        htmltools::span(
            htmltools::span(class = "legend-badge", style = "background-color: rgba(74,127,181,0.5);"),
            "Literature occurrence"
        ),
        htmltools::span(
            htmltools::span(class = "legend-badge", style = "background-color: rgba(248,250,252,0.6); border: 1px solid #e2e8f0;"),
            "No occurrence"
        )
    ),
    htmltools::tags$a(
        href = "Wilkens et al - Fossil Dataset - 12-02-26.xlsx",
        class = "btn-download",
        download = "Wilkens et al - Fossil Dataset - 12-02-26.xlsx",
        htmltools::tags$i(class = "fa fa-download"),
        "Download"
    )
)

dt <- datatable(
    dashboard_data_display,
    filter = "none",
    rownames = FALSE,
    class = "stripe hover compact cell-border",
    options = list(
        pageLength = -1,
        lengthMenu = list(c(25, 50, 100, -1), c("25", "50", "100", "All")),
        scrollY = "calc(100vh - 340px)",
        scrollX = TRUE,
        scrollCollapse = TRUE,
        autoWidth = FALSE,
        columnDefs = list(
            list(width = "75px", targets = as.list(
                which(names(dashboard_data_display) %in% c("genus", "family", period_cols)) - 1L
            ))
        ),
        dom = "frti",
        language = list(
            search      = "Search genera:",
            info        = "Showing _START_ to _END_ of _TOTAL_ genera",
            infoEmpty   = "No genera found",
            zeroRecords = "No matching genera",
            lengthMenu  = "Show _MENU_ genera"
        ),
        headerCallback = JS(sprintf("
            function(thead, data, start, end, display) {
                var colColors = %s;
                $(thead).find('th').each(function(i) {
                    if (colColors[i] !== undefined) {
                        $(this).css({'background-color': colColors[i], 'color': '#1e293b'});
                    }
                });
            }
        ", {
            idx <- which(names(dashboard_data_display) %in% names(colors_periods)) - 1L
            nms <- names(dashboard_data_display)[idx + 1L]
            pairs <- paste(sprintf('"%d":"%s"', idx, colors_periods[nms]), collapse = ",")
            paste0("{", pairs, "}")
        })),
        initComplete = JS("
            function(settings, json) {
                var api = this.api();
                $(api.table().container())
                    .css({'font-size': '12px', 'font-family': 'Inter, sans-serif'});
                $(api.table().header())
                    .css({'font-size': '13px', 'font-family': 'Inter, sans-serif'});
                setTimeout(function() { api.columns.adjust(); }, 10);
            }
        ")
    )
) %>%
    formatStyle(
        columns = period_cols,
        backgroundColor = styleEqual(
            c("0", "1", "2"),
            c("rgba(248,250,252,0.6)", "rgba(74,127,181,0.5)", "rgba(30,58,95,0.55)")
        ),
        color = styleEqual(
            c("0", "1", "2"),
            c("rgba(248,250,252,0.6)", "rgba(74,127,181,0.5)", "rgba(30,58,95,0.55)")
        ),
        fontSize = "1px",
        textAlign = "center",
        border = "1px solid #e2e8f0"
    ) %>%
    formatStyle(
        columns    = "genus",
        fontStyle  = "italic",
        fontWeight = "600",
        color      = "#1e293b"
    ) %>%
    formatStyle(
        columns    = "family",
        color      = "#64748b",
        fontSize   = "10px"
    ) %>%
    formatStyle(
        columns = "notes",
        fontSize = "10px",
        color = "#475569",
        maxWidth = "300px",
        overflow = "hidden",
        textOverflow = "ellipsis",
        whiteSpace = "nowrap",
        cursor = "pointer"
    ) %>%
    formatStyle(
        columns = "references",
        fontSize = "10px",
        color = "#475569",
        maxWidth = "250px",
        overflow = "hidden",
        textOverflow = "ellipsis",
        whiteSpace = "nowrap",
        cursor = "pointer"
    )

# JavaScript to toggle cell expansion on click
expand_js <- htmltools::tags$script(htmltools::HTML("
$(document).ready(function() {
    $(document).on('click', 'td', function() {
        var $td = $(this);
        // Only apply to notes/references columns (check by index or content length)
        if ($td.css('white-space') === 'nowrap' && $td.css('cursor') === 'pointer') {
            if ($td.data('expanded')) {
                $td.css({'white-space': 'nowrap', 'max-width': $td.data('orig-width'), 'overflow': 'hidden', 'text-overflow': 'ellipsis'});
                $td.data('expanded', false);
            } else {
                $td.data('orig-width', $td.css('max-width'));
                $td.css({'white-space': 'normal', 'max-width': '500px', 'overflow': 'visible', 'text-overflow': 'clip'});
                $td.data('expanded', true);
            }
        }
    });
});
"))

htmltools::tagList(header_html, dt, expand_js)
```

<script>
function ptopToggle(which) {
    var gd = document.getElementById('p-top-climate');
    if (!gd) return;
    var vis = which === 'upper'
        ? [false, false, false, true,  true,  true]
        : [true,  true,  true,  false, false, false];
    Plotly.restyle(gd, {visible: vis});
    document.getElementById('ptop-btn-lower').style.background = which === 'lower' ? '#cbd5e1' : '#f1f5f9';
    document.getElementById('ptop-btn-upper').style.background = which === 'upper' ? '#cbd5e1' : '#f1f5f9';
}
</script>
