---
title: "Prepare data for niche centers and create plots for Fig. 4"
output:
  html_document:
    df_print: paged
---

Load in necessary libraries.
```{r, eval = FALSE}
library(brms)
library(dplyr)
library(ggplot2)
library(here)
library(terra)
library(patchwork)
library(readxl)
```

Load in fossil data, global tree genus checklist, and native species/genus checklists for WCE.
```{r}
dataset_final <- read_xlsx(here("./data/data_1/dataset_final_12_02_26.xlsx"))
native_genera <- read.csv(here("./data/data_1/native_genera.csv"))[, -1]
native_species <- read.csv(here("./data/data_1/native_species.csv"))[, -1]
trees_final <- read.csv(here("./data/data_1/trees_final.csv"))


mm <- function(x) {
  x / 0.75
}
```

Load in climate data for WCE and reproject into equal-area projection EPSG:3035.
```{r}
bioclim_current <- rast(here("./data/data_2/bioclim_current_0_25deg.tif"))
bioclim_ssp126 <- rast(here("./data/data_2/bioclim_ssp126_0_25deg.tif"))
bioclim_ssp585 <- rast(here("./data/data_2/bioclim_ssp585_0_25deg.tif"))

bioclim_current <- project(bioclim_current, "EPSG:3035")
bioclim_ssp126 <- project(bioclim_ssp126, "EPSG:3035")
bioclim_ssp585 <- project(bioclim_ssp585, "EPSG:3035")

bioclim_current_df <- as.data.frame(bioclim_current, xy = TRUE)
colnames(bioclim_current_df) <- c("x", "y", "bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19")

bioclim_ssp126_df <- as.data.frame(bioclim_ssp126, xy = TRUE)
colnames(bioclim_ssp126_df) <- c("x", "y", "bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19")

bioclim_ssp585_df <- as.data.frame(bioclim_ssp585, xy = TRUE)
colnames(bioclim_ssp585_df) <- c("x", "y", "bio1", "bio2", "bio3", "bio4", "bio5", "bio6", "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", "bio19")

bioclim_current_df$group <- "current"
bioclim_ssp126_df$group <- "ssp126"
bioclim_ssp585_df$group <- "ssp585"

combined_data <- rbind(bioclim_current_df, bioclim_ssp126_df, bioclim_ssp585_df)
combined_data <- na.omit(combined_data)
```

Prepare data for Fig. 4 by loading in all models of niche centers and extracting desired information, including the genus-level fixed intercepts, Rhat, number of divergent transitions, and number of species used to fit the models.
```{r}
# Input directory
directory_path <- here("./data/niche_centers")

file_paths <- list.files(directory_path, full.names = TRUE)
model_list <- lapply(file_paths, readRDS)

model_names <- sub(".*model_(.*)\\.RData", "\\1", file_paths)
genus_names <- sub("_na$", "", model_names)

# Define function to extract the number of divergent transitions from each model object.
n_divergent <- function(x) {
  stopifnot(is(x, "brmsfit"))
  out <- lapply(x$fit@sim$samples, function(y) {
    sum(attr(y, "sampler_params")[["divergent__"]])
  })
  sum(unlist(out))
}

# Initialize dataframe
all_data_centers <- data.frame()

# Loop through all models by loading in the model object and extracting desired information.
for (g in 1:length(genus_names)) {
  genus_name <- genus_names[[g]]
  model_name <- model_names[[g]]
  model <- model_list[[g]]
  Rhat <- max(rhat(model))
  species_n <- length(unique(model$data$species))
  divtrans <- n_divergent(model)

  # If the model in question is monospecific, the format is different:
  if (ncol(as_draws_df(model)) == 11) {
    para <- as_draws_df(model)
    para <- select(para, b_bio1_Intercept, b_bio12_Intercept)
    para <- colMeans(para)
    df <- data.frame(genus = genus_name, model = model_name, bio1 = para[1], bio12 = para[2], Rhat = Rhat, n = species_n, divtrans = divtrans)
    all_data_centers <- rbind(all_data_centers, df)
    next
  }

  para <- as_draws_df(model)
  para <- select(para, b_bio1_Intercept, b_bio12_Intercept)
  para <- colMeans(para)
  df <- data.frame(genus = genus_name, model = model_name, bio1 = para[1], bio12 = para[2], Rhat = Rhat, n = species_n, divtrans = divtrans)

  all_data_centers <- rbind(all_data_centers, df)
}

all_data_centers$category <- ifelse(all_data_centers$genus %in% native_species | grepl("_na$", all_data_centers$model), "HRTF",
  ifelse(all_data_centers$genus %in% native_genera, "EG", "REG")
)


all_data_centers$n <- ifelse(all_data_centers$n == 0, 1, all_data_centers$n)

all_data_centers$plot_Rhat <- ifelse(all_data_centers$Rhat <= 1.05, 1, 0)
all_data_centers$plot_divtrans <- ifelse(all_data_centers$divtrans <= 40, 1, 0)

paleotrees_adj <- dataset_final
paleotrees_adj$last_presence <- apply(paleotrees_adj[, -1], 1, function(x) {
  if (any(x == 1)) names(x)[max(which(x == 1))] else NA
})

all_data_centers <- merge(all_data_centers, paleotrees_adj[, c(1, 13)], by = "genus")
all_data_centers$last_presence <- ifelse(all_data_centers$last_presence %in% c("Eem", "Tegelen", "Waal", "Cromer", "Holstein"), "Pleistocene", all_data_centers$last_presence)

all_data_centers$label <- all_data_centers$genus
all_data_centers$label <- paste0("italic('", all_data_centers$label, "')")

all_data_centers$label <- with(all_data_centers, paste0(
  label, "[",
  ifelse(category == "HRTF", "\"nat.\"", ""), "]"
))

total_species <- data.frame()
for (g in 1:length(genus_names)) {
  genus_name <- genus_names[[g]]
  model_name <- model_names[[g]]

  if (genus_name != model_name) {
    species_list <- filter(trees_final, trees_final$genus == genus_name)
    species_list <- filter(species_list, species_list$species %in% native_species)
    df <- data.frame(model = model_name, n_total = nrow(species_list))
    total_species <- rbind(total_species, df)
  }

  if (genus_name == model_name) {
    species_list <- filter(trees_final, trees_final$genus == genus_name)
    df <- data.frame(model = model_name, n_total = nrow(species_list))
    total_species <- rbind(total_species, df)
  }
}

all_data_centers <- merge(all_data_centers, total_species, by = "model")
```

Create the plot for Fig. 4a:
```{r}
# Define a list of genera whose labels will be plotted
plot_labels <- c("Picea_na", "Picea", "Cedrus", "Pseudotsuga", "Aesculus", "Juglans", "Castanea", "Phoenix", "Ziziphus", "Taiwania", "Tsuga", "Areca", "Meliosma", "Eugenia", "Tetraclinis", "Hippophae_na", "Hippophae", "Fagus", "Fagus_na", "Picea", "Picea_na", "Acer", "Acer_na", "Laurus", "Ficus")
all_data_centers$plot_label <- ifelse(all_data_centers$model %in% plot_labels, 1, 0)


p1 <- ggplot() +
  stat_density_2d(
    data = combined_data,
    aes(x = bio1, y = bio12, fill = group, alpha = after_stat(level)),
    geom = "polygon",
    color = NA,
    contour_var = "ndensity",
    bins = 7
  ) +
  geom_point(
    data = all_data_centers[
      all_data_centers$plot_Rhat == 1 &
        all_data_centers$plot_divtrans == 1,
    ],
    aes(x = bio1, y = bio12, fill = category, size = n_total, shape = category),
    color = "black",
    stroke = 0.25,
    show.legend = FALSE
  ) +

  # Invisible layers used only to construct the combined size legend
  geom_point(
    data = data.frame(bio1 = 3, bio12 = 400, n_total = 1),
    aes(x = bio1, y = bio12, size = n_total),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.25,
    alpha = 0,
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +
  geom_point(
    data = data.frame(bio1 = 3, bio12 = 400, n_total = 1),
    aes(x = bio1, y = bio12, size = n_total),
    shape = 24,
    fill = "black",
    color = "black",
    stroke = 0.25,
    alpha = 0,
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +

  scale_x_continuous(
    breaks = c(5, 10, 15, 20, 25, 30),
    limits = c(3, 28),
    minor_breaks = setdiff(
      seq(3, 28, by = 1),
      c(5, 10, 15, 20, 25, 30)
    ),
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  scale_y_continuous(
    limits = c(400, 3200),
    breaks = c(500, 1000, 1500, 2000, 2500, 3000),
    minor_breaks = setdiff(
      seq(400, 3200, by = 100),
      c(500, 1000, 1500, 2000, 2500, 3000)
    ),
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  geom_text(
    data = all_data_centers[all_data_centers$plot_label == 1, ],
    aes(x = bio1, y = bio12, label = label),
    parse = TRUE,
    size = 6 / .pt
  ) +
  stat_ellipse(
    data = all_data_centers[all_data_centers$category == "HRTF", ],
    aes(x = bio1, y = bio12),
    color = "black",
    linetype = "longdash",
    linewidth = mm(0.15)
  ) +
  scale_color_manual(
    values = c(
      "current" = "#909090",
      "EG" = "#FFD579",
      "HRTF" = "#237B55",
      "REG" = "#69A0C9",
      "ssp126" = "#FF9538",
      "ssp585" = "#EB0000"
    )
  ) +
  scale_fill_manual(
    values = c(
      "current" = "#909090",
      "EG" = "#FFD579",
      "HRTF" = "#237B55",
      "REG" = "#69A0C9",
      "ssp126" = "#FF9538",
      "ssp585" = "#EB0000"
    )
  ) +
  scale_size_continuous(
    name = "Number of records",
    range = c(1, 10),
    limits = c(1, 850),
    breaks = c(1, 10, 100, 500, 850)
  ) +
  scale_linetype_manual(values = c("longdash", "solid", "dotted")) +
  scale_shape_manual(
    values = c(
      "EG" = 21,
      "HRTF" = 24,
      "REG" = 21
    )
  ) +
  guides(
    fill = "none",
    shape = "none",
    alpha = "none",
    color = "none",
    linetype = "none",
    size = guide_legend(
      title = "Number of records",
      reverse = TRUE,
      override.aes = list(alpha = 1)
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.72),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.8),
      color = NA
    ),
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "black", linewidth = mm(0.15)),
    axis.line.y = element_line(colour = "black", linewidth = mm(0.15)),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.text.y = element_text(size = 8, colour = "black"),
    axis.title.y = element_blank(),
    axis.ticks.length = unit(0.1, "cm"),
    axis.ticks.x = element_line(colour = "black", linewidth = mm(0.15)),
    axis.ticks.y = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.y.left = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.length.y = rel(0.75),
    axis.minor.ticks.x.bottom = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.length.x = rel(0.75)
  )

ggsave(plot = p1, filename = "Fig4a_niche_centers.pdf", path = here("figures"), width = 140, height = 140, units = "mm", device = "pdf", bg = "white")

```

Create the plot for Fig. 4b:

```{r}

colors_periods <- c(
  "Holocene" = "#DCE8E9",
  "Eem" = "#CBF5FC",
  "Holstein" = "#B4E2E3",
  "Cromer" = "#99D9DD",
  "Waal" = "#7FC5DD",
  "Tegelen" = "#4FB3D4",
  "Pliocene" = "#FFF6B2",
  "Miocene" = "#FFEB3D",
  "Oligocene" = "#FFC694",
  "Eocene" = "#FFBC87",
  "Pleistocene" = "#99d9ddff"
)



p2 <- ggplot() +
  geom_point(
    data = all_data_centers[
      all_data_centers$plot_Rhat == 1 &
        all_data_centers$plot_divtrans == 1,
    ],
    aes(x = bio1, y = bio12, fill = last_presence, size = n_total, shape = category),
    color = "black",
    stroke = 0.25,
    show.legend = FALSE
  ) +

  # Invisible layers used only to construct the combined size legend
  geom_point(
    data = data.frame(bio1 = 3, bio12 = 400, n_total = 1),
    aes(x = bio1, y = bio12, size = n_total),
    shape = 21,
    fill = "white",
    color = "black",
    stroke = 0.25,
    alpha = 0,
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +
  geom_point(
    data = data.frame(bio1 = 3, bio12 = 400, n_total = 1),
    aes(x = bio1, y = bio12, size = n_total),
    shape = 24,
    fill = "black",
    color = "black",
    stroke = 0.25,
    alpha = 0,
    inherit.aes = FALSE,
    show.legend = TRUE
  ) +

  scale_fill_manual(values = colors_periods) +
  scale_size_continuous(
    name = "Number of records",
    range = c(1, 5),
    limits = c(1, 850),
    breaks = c(1, 10, 100, 500, 850)
  ) +
  stat_ellipse(
    data = all_data_centers[all_data_centers$category == "HRTF", ],
    aes(x = bio1, y = bio12),
    color = "black",
    linetype = "longdash",
    linewidth = mm(0.15)
  ) +
  scale_shape_manual(
    values = c(
      "EG" = 21,
      "HRTF" = 24,
      "REG" = 21
    )
  ) +
  scale_x_continuous(
    breaks = c(5, 10, 15, 20, 25, 30),
    limits = c(3, 28),
    minor_breaks = setdiff(
      seq(3, 28, by = 1),
      c(5, 10, 15, 20, 25, 30)
    ),
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  scale_y_continuous(
    limits = c(400, 3200),
    breaks = c(500, 1000, 1500, 2000, 2500, 3000),
    minor_breaks = setdiff(
      seq(400, 3200, by = 100),
      c(500, 1000, 1500, 2000, 2500, 3000)
    ),
    guide = guide_axis(minor.ticks = TRUE)
  ) +
  guides(
    fill = "none",
    shape = "none",
    alpha = "none",
    color = "none",
    linetype = "none",
    size = guide_legend(
      title = "Number of records",
      reverse = TRUE,
      override.aes = list(alpha = 1)
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "inside",
    legend.position.inside = c(0.85, 0.72),
    legend.background = element_rect(
      fill = scales::alpha("white", 0.8),
      color = NA
    ),
    panel.background = element_blank(),
    panel.grid.major = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(colour = "black", linewidth = mm(0.15)),
    axis.line.y = element_line(colour = "black", linewidth = mm(0.15)),
    axis.text.x = element_text(size = 8, colour = "black"),
    axis.text.y = element_text(size = 8, colour = "black"),
    axis.title.y = element_blank(),
    axis.ticks.length = unit(0.1, "cm"),
    axis.ticks.x = element_line(colour = "black", linewidth = mm(0.15)),
    axis.ticks.y = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.y.left = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.length.y = rel(0.75),
    axis.minor.ticks.x.bottom = element_line(colour = "black", linewidth = mm(0.15)),
    axis.minor.ticks.length.x = rel(0.75)
  )

ggsave(plot = p2, filename = "Fig4b_last_occurrences.pdf", path = here("figures"), width = 140, height = 60, device = "pdf", units = "mm", bg = "white")

```